A Framework for Substructural Type Systems \thanks{James Wood is supported by an EPSRC Studentship. Robert Atkey is supported by EPSRC grant FIXME.}

### 2022-01-25 tex/esop22/examples.tex
[ esop22 ] Reorder Section 6, try to smooth narrative

>>>>

\bob{Reification dumped here for now} When \AgdaBound{$\V$} supports identity environments (as per \cref{thm:env-id}),
we can coerce a \AgdaFunction{Kripke} function into just a
\AgdaBound{$\C$}-value in an extended context.
To show this, we instantiate the Kripke function with the renaming
$\swarrow^r : \Gamma, \gr0\delta \env\sqni \Gamma$ to extend the scope, and
pass it an argument environment
$\plr{\id^\V; \searrow^r} : \gr0\gamma, \Delta \env\V \Delta$ to fill in the
extended part.
Post-composition of a renaming onto an environment comes from
\cref{thm:env-postren}.

\ExecuteMetaData[\Syntactictex]{reify}

We tend to use \AgdaFunction{reify} whenever \AgdaBound{$\V$} supports
identity environments.
In this paper, all examples support identity environments.
However, \citet[p.~27]{AACMM21} give some important examples that do not
support identity environments, and thus cannot use \AgdaFunction{reify}.
The feature that causes the lack of support for identity environments is that
a semantics can make use of the fact that only variables of particular kinds
are bound by the syntax.
In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
binds variables that synthesise their type, as opposed to those whose type is
checked.
The semantics of type-checking and elaboration rely on variables synthesising
their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
only binds such synthesising variables.
Similar phenomena would appear in, say, a call-by-value language where
variables are values (not computations), or a polarised language where
variables always have a polarity matching their type.

### 2022-01-25 tex/esop22/semantics.tex
[ esop22 ] Reorder Section 6, try to smooth narrative

Our aim with the following traversals is to derive a recursion scheme for
terms which abstracts over some of the details of variable binding and
changes to usage annotations.
We follow the structure given by \citet{AACMM21} closely, with
the main difference having come already in the syntax and environments.

@@

\begin{lemma}[\AgdaFunction{ren\textasciicircum{}Env}]\label{thm:env-ren}
  If $\plr{-} \sdtstile{}\V A$ is renameable for all $A$, then so is
  $\plr{-} \env\V \Delta$ for all $\Delta$.
\end{lemma}
\begin{proof}
  For renameable $\V$, we can implement $\mathrm{lift}$ as in
  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
  $\V \coloneqq \W \coloneqq \V$.
  Renaming the environment then becomes an instance of environment composition.
\end{proof}

\subsection{The Kripke function space}

Earlier, when talking about syntax, we saw \AgdaFunction{Scope} as a simple way
to let an open family admit context extensions.
However, a key component of our generic semantic traversal is to make use of
the open family \AgdaBound{$\V$} of \emph{values}, which are the sort of thing
we store in an environment.
The definition \AgdaFunction{Kripke} gives an alternative to
\AgdaFunction{Scope} which interprets the newly bound variables via a
requirement of $\V$-values rather than extra assumptions for the
$\C$-computation.

When \AgdaBound{$\V$} supports identity environments (as per \cref{thm:env-id}),
we can coerce a \AgdaFunction{Kripke} function into just a
\AgdaBound{$\C$}-value in an extended context.
To show this, we instantiate the Kripke function with the renaming
$\swarrow^r : \Gamma, \gr0\delta \env\sqni \Gamma$ to extend the scope, and
pass it an argument environment
$\plr{\id^\V; \searrow^r} : \gr0\gamma, \Delta \env\V \Delta$ to fill in the
extended part.
Post-composition of a renaming onto an environment comes from
\cref{thm:env-postren}.

\ExecuteMetaData[\Syntactictex]{reify}

We tend to use \AgdaFunction{reify} whenever \AgdaBound{$\V$} supports
identity environments.
In this paper, all examples support identity environments.
However, \citet[p.~27]{AACMM21} give some important examples that do not
support identity environments, and thus cannot use \AgdaFunction{reify}.
The feature that causes the lack of support for identity environments is that
a semantics can make use of the fact that only variables of particular kinds
are bound by the syntax.
In the examples of \citeauthor{AACMM21}, a bidirectionally typed language only
binds variables that synthesise their type, as opposed to those whose type is
checked.
The semantics of type-checking and elaboration rely on variables synthesising
their type, so \AgdaBound{$\V$} is chosen to cover only those variables.
Instead of using \AgdaFunction{reify}, we must observe that each syntactic form
only binds such synthesising variables.
Similar phenomena would appear in, say, a call-by-value language where
variables are values (not computations), or a polarised language where
variables always have a polarity matching their type.

\subsection{A layer of syntax is functorial}

A basic property of the universe of syntaxes we described in \cref{sec:syntax}
is that every syntax supports a functorial action on subterms.
The type of \AgdaFunction{map-s} says that to map a function \AgdaBound{f}
over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
domain and codomain usage contexts, and \AgdaBound{f} should be usable
wherever the domain and codomain usage contexts are similarly related by
\AgdaBound{F}.
This generality is needed because usage contexts typically will change between
a term and its immediate subterms, but there will be a linear combination of
usage contexts for the subterms that make the original usage contexts.
\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
variables newly bound in that subterm).
Unlike usage annotations, types are preserved throughout a term (including
\AgdaBound{$\gamma$}, \AgdaBound{$\delta$}, and the implicit type of the term).

\subsection{Semantic traversal}

>>>>

Our goal in this section is to define \AgdaFunction{semantics}, a
recursor that turns a term into a \AgdaBound{$\C$}-value using a
\AgdaBound{$\V$}-environment, in a type preserving way:\bob{Get rid of
  ``body'' here}

The \AgdaBound{$\V$} and \AgdaBound{$\C$} are \AgdaFunction{OpenFam}s,
representing the interpretations of variables and terms
respectively. In \cref{sec:traversal} we will see the data that must
be provided to make a \AgdaFunction{semantics} for a given
system. Before that, we must see how to deal with the two complicated
features of our syntax: the usage annotations (\cref{sec:functorial})
and variable binding (\cref{sec:kripke}). \todo{fwd ref to where these are used}

\subsection{A layer of syntax is functorial}\label{sec:functorial}

A basic property of the universe of syntaxes we described in \cref{sec:syntax}
is that every syntax supports a functorial action on subterms, realised by the function \AgdaFunction{map-s}.
Its says that to map a function \AgdaBound{f}
over a layer of syntax, there must be a linear map \AgdaBound{F} relating the
domain and codomain usage contexts, and \AgdaBound{f} should be usable
wherever the domain and codomain usage contexts are similarly related by
\AgdaBound{F}.

This generality is needed because usage contexts change between
a term and its immediate subterms, but there will be a linear combination of
the subterms' usage contexts for the subterms that make the original usage contexts.
\AgdaBound{X} and \AgdaBound{Y} are \AgdaFunction{ExtOpenFam}s, with
\AgdaBound{$\Theta$} being the context extension for a subterm (i.e., the
variables newly bound in that subterm).
Unlike usage annotations, types are preserved throughout a term (including
\AgdaBound{$\gamma$}, \AgdaBound{$\delta$}, and the implicit type of the term).

\subsection{The Kripke function space}\label{sec:kripke}

In \secref{sec:terms}, we defined \AgdaFunction{Scope} to let a
judgement-indexed family admit context extensions. However, a key
component of our generic semantic traversal is to make use of the open
family \AgdaBound{$\V$} of \emph{values}, which are the sort of thing
we store in an environment.  The definition \AgdaFunction{Kripke}
gives an alternative to \AgdaFunction{Scope} which interprets the
newly bound variables via a requirement of $\V$-values rather than
extra assumptions for the $\C$-computation.

@@

\begin{lemma}[\AgdaFunction{ren\textasciicircum{}Env}]\label{thm:env-ren}
  If $\plr{-} \sdtstile{}\V A$ is renameable for all $A$, then so is
  $\plr{-} \env\V \Delta$ for all $\Delta$.
\end{lemma}
\bob{Proof probably isn't needed}
\begin{proof}
  For renameable $\V$, we can implement $\mathrm{lift}$ as in
  \cref{thm:env-comp} with $\U \coloneqq {\sqni}$ and
  $\V \coloneqq \W \coloneqq \V$.
  Renaming the environment then becomes an instance of environment composition.
\end{proof}

\subsection{Semantic traversal}\label{sec:traversal}
